---
title: "Poem Embeddings Explorer"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    self-contained: true
---

> **Requires:** An OpenAI API key stored as `OPENAI_API_KEY` in your `.Renviron`
> (run `usethis::edit_r_environ()` to open it). You'll also need
> `install.packages(c("openai", "ggrepel"))` if not already installed.

---

## Configuration

This is the only section you need to edit when switching poems. All downstream
chunks read from the objects defined here.

```{r}
#| label: config

library(tidyverse)
library(openai)

# ── 1. Poem metadata ──────────────────────────────────────────────────────────
poem_title  <- "Digging"
poem_author <- "Seamus Heaney"

# ── 2. Poem text ──────────────────────────────────────────────────────────────
# Paste the full poem as a plain string.
# Separate stanzas with a blank line; the parser does the rest.
poem_raw <- "
Between my finger and my thumb
The squat pen rests; snug as a gun.

Under my window, a clean rasping sound
When the spade sinks into gravelly ground:
My father, digging. I look down

Till his straining rump among the flowerbeds
Bends low, comes up twenty years away
Stooping in rhythm through potato drills
Where he was digging.

The coarse boot nestled on the lug, the shaft
Against the inside knee was levered firmly.
He rooted out tall tops, buried the bright edge deep
To scatter new potatoes that we picked,
Loving their cool hardness in our hands.

By God, the old man could handle a spade.
Just like his old man.

My grandfather cut more turf in a day
Than any other man on Toner's bog.
Once I carried him milk in a bottle
Corked sloppily with paper. He straightened up
To drink it, then fell to right away
Nicking and slicing neatly, heaving sods
Over his shoulder, going down and down
For the good turf. Digging.

The cold smell of potato mould, the squelch and slap
Of soggy peat, the curt cuts of an edge
Through living roots awaken in my head.
But I've no spade to follow men like them.

Between my finger and my thumb
The squat pen rests.
I'll dig with it.
"

# ── 3. Stanza arc reference points ────────────────────────────────────────────
# Which stanzas to compare all others against in the arc plot?
# Use 1-based integer indices. NA resolves to the last stanza.
arc_refs <- c(1L, NA)

# ── 4. Concept probes ─────────────────────────────────────────────────────────
# Named character vector: the name is the display label, the value is the
# phrase sent to the embedding model. Specific phrases work better than
# single words — give the model enough context to anchor the concept.
concept_probes <- c(
  "physical labour and toil" = "physical labour and toil",
  "memory and the past"      = "memory and the past",
  "writing and the pen"      = "writing and the pen",
  "nature and the earth"     = "nature and the earth",
  "family and ancestry"      = "family and ancestry"
)

# ── 5. Topic keywords (whole-poem scoring) ────────────────────────────────────
# A plain character vector of keywords or short phrases. These are used to
# produce a single topic-score vector for the whole poem — useful for
# comparing poems against each other later.
# Single concrete words work fine; abstract concepts benefit from a short phrase.
topic_keywords <- c("memory", "labor", "father", "legacy")
```

---

## Parse

```{r}
#| label: parse
#| dependson: "config"

# Split on blank lines → stanzas; split stanzas on newlines → lines
stanza_texts <- str_split(str_trim(poem_raw), "\n\\s*\n")[[1]] |>
  str_trim() |>
  keep(\(x) nchar(x) > 0)

poem_stanzas <- tibble(
  stanza_num = seq_along(stanza_texts),
  # Single-string version for embedding (newlines → spaces)
  text       = str_replace_all(stanza_texts, "\\s*\n\\s*", " "),
  # Short label: stanza number + first few words
  label      = paste0("S", stanza_num, ": ",
                      str_trunc(str_extract(stanza_texts, "^[^\n]+"), 22))
)

poem_lines <- map(seq_along(stanza_texts), \(i) {
  lines <- str_split(stanza_texts[i], "\n")[[1]] |>
    str_trim() |>
    keep(\(x) nchar(x) > 0)
  tibble(stanza = i, text = lines)
}) |>
  list_rbind() |>
  mutate(line_num = row_number())

# Resolve NA in arc_refs to last stanza index
arc_refs <- ifelse(is.na(arc_refs), nrow(poem_stanzas), arc_refs)

cat(glue::glue(
  '"{poem_title}" by {poem_author}\n',
  '{nrow(poem_lines)} lines across {nrow(poem_stanzas)} stanzas'
))
```

---

## The Poem

```{r}
#| label: display
#| echo: false
#| results: asis
#| dependson: "parse"

cat(sprintf(
  '::: {style="font-family: Georgia, serif; font-size: 1.05em; line-height: 1.9; max-width: 480px;"}\n\n### %s\n\n*%s*\n\n',
  poem_title, poem_author
))

for (st in stanza_texts) {
  lines <- str_split(st, "\n")[[1]] |> str_trim()
  cat(paste(lines, collapse = "  \n"))
  cat("\n\n")
}

cat(":::\n")
```

---

## Embeddings

```{r}
#| label: embeddings
#| cache: true
#| dependson: "parse"

embed_texts <- function(texts, model = "text-embedding-3-small") {
  result <- openai::create_embedding(model = model, input = texts)
  do.call(rbind, result$data$embedding)
}

line_emb   <- embed_texts(poem_lines$text)
stanza_emb <- embed_texts(poem_stanzas$text)
```

---

## Cosine Similarity

Cosine similarity measures the angle between two embedding vectors: 1 means
identical direction (very similar meaning), 0 means orthogonal (unrelated).

```{r}
#| label: cosine-fn
#| dependson: "embeddings"

cosine_matrix <- function(mat) {
  norms    <- sqrt(rowSums(mat^2))
  mat_norm <- mat / norms
  round(mat_norm %*% t(mat_norm), 3)
}

line_sim   <- cosine_matrix(line_emb)
stanza_sim <- cosine_matrix(stanza_emb)
```

### Line-level heatmap

Each cell shows how semantically similar two lines are. The diagonal is always
1 (a line compared to itself).

```{r}
#| label: line-heatmap
#| fig-width: 9
#| fig-height: 8
#| dependson: "cosine-fn"

short_label <- paste0(poem_lines$line_num, ": ",
                      str_trunc(poem_lines$text, 26))

expand.grid(line_a = seq_len(nrow(line_sim)),
            line_b = seq_len(nrow(line_sim))) |>
  mutate(
    similarity = line_sim[cbind(line_a, line_b)],
    label_a    = factor(short_label[line_a], levels = short_label),
    label_b    = factor(short_label[line_b], levels = rev(short_label))
  ) |>
  ggplot(aes(x = label_a, y = label_b, fill = similarity)) +
  geom_tile() +
  scale_fill_distiller(palette = "Blues", direction = 1,
                       name = "Cosine\nsimilarity") +
  theme_minimal(base_size = 8) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title  = element_blank(),
    panel.grid  = element_blank()
  ) +
  labs(title = glue::glue("Line-level cosine similarity — {poem_title}"))
```

### Stanza-level heatmap

```{r}
#| label: stanza-heatmap
#| fig-width: 7
#| fig-height: 6
#| dependson: "cosine-fn"

expand.grid(stanza_a = seq_len(nrow(stanza_sim)),
            stanza_b = seq_len(nrow(stanza_sim))) |>
  mutate(
    similarity = stanza_sim[cbind(stanza_a, stanza_b)],
    label_a    = factor(poem_stanzas$label[stanza_a],
                        levels = poem_stanzas$label),
    label_b    = factor(poem_stanzas$label[stanza_b],
                        levels = rev(poem_stanzas$label))
  ) |>
  ggplot(aes(x = label_a, y = label_b, fill = similarity)) +
  geom_tile() +
  geom_text(aes(label = round(similarity, 2)), size = 3.2) +
  scale_fill_distiller(palette = "Blues", direction = 1,
                       name = "Cosine\nsimilarity") +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1),
    axis.title  = element_blank(),
    panel.grid  = element_blank()
  ) +
  labs(title = glue::glue("Stanza-level cosine similarity — {poem_title}"))
```

---

## PCA: Lines in Embedding Space

```{r}
#| label: pca-lines
#| fig-width: 9
#| fig-height: 7
#| dependson: "embeddings"

pca_fit <- prcomp(line_emb, scale. = FALSE)
var_exp <- summary(pca_fit)$importance["Proportion of Variance", 1:2] * 100

as_tibble(pca_fit$x[, 1:2]) |>
  bind_cols(poem_lines) |>
  mutate(short_text = str_trunc(text, 32)) |>
  ggplot(aes(x = PC1, y = PC2, color = factor(stanza), label = short_text)) +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_text_repel(size = 2.8, max.overlaps = 20,
                            show.legend = FALSE) +
  scale_color_brewer(palette = "Set2", name = "Stanza") +
  labs(
    title = glue::glue("Lines in PCA embedding space — {poem_title}"),
    x     = sprintf("PC1 (%.1f%% variance)", var_exp[1]),
    y     = sprintf("PC2 (%.1f%% variance)", var_exp[2])
  ) +
  theme_minimal(base_size = 11)
```

---

## Most & Least Similar Line Pairs

```{r}
#| label: top-pairs
#| dependson: "cosine-fn"

line_sim_pairs <- expand.grid(
    line_a = seq_len(nrow(line_sim)),
    line_b = seq_len(nrow(line_sim))
  ) |>
  filter(line_a < line_b) |>
  mutate(
    similarity = line_sim[cbind(line_a, line_b)],
    text_a     = poem_lines$text[line_a],
    text_b     = poem_lines$text[line_b]
  )

cat("--- Most similar pairs ---\n")
line_sim_pairs |>
  slice_max(similarity, n = 8) |>
  select(similarity, text_a, text_b) |>
  print()

cat("\n--- Least similar pairs ---\n")
line_sim_pairs |>
  slice_min(similarity, n = 8) |>
  select(similarity, text_a, text_b) |>
  print()
```

---

## Stanza Arc

How does each stanza relate semantically to the reference stanzas set in
`arc_refs`? This is most revealing when the reference stanzas represent
structural anchors — an opening image, a volta, a closing resolution.

```{r}
#| label: stanza-arc
#| fig-width: 8
#| fig-height: 4
#| dependson: "cosine-fn"

map(arc_refs, \(ref) {
  tibble(
    stanza_num = poem_stanzas$stanza_num,
    label      = poem_stanzas$label,
    reference  = paste0("Ref S", ref, ": ",
                        str_trunc(poem_stanzas$label[ref], 25)),
    similarity = stanza_sim[, ref]
  )
}) |>
  list_rbind() |>
  ggplot(aes(x = stanza_num, y = similarity,
             color = reference, group = reference)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = poem_stanzas$stanza_num,
                     labels = poem_stanzas$label,
                     guide  = guide_axis(angle = 35)) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(
    title = glue::glue("Stanza similarity to reference points — {poem_title}"),
    x     = NULL,
    y     = "Cosine similarity"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

## Concept Probing

Individual embedding dimensions are not human-interpretable — meaning is
distributed across all dimensions simultaneously. But we can define semantic
directions by embedding a concept phrase and projecting each line onto that
direction. Edit `concept_probes` in the **Configuration** section at the top
to change or add concepts.

```{r}
#| label: concept-probing
#| cache: true
#| fig-width: 9
#| fig-height: 7
#| dependson: ["embeddings", "config"]

concept_result <- openai::create_embedding(
  model = "text-embedding-3-small",
  input = unname(concept_probes)
)
concept_mat <- do.call(rbind, concept_result$data$embedding)

l2_norm <- function(mat) mat / sqrt(rowSums(mat^2))

scores <- l2_norm(line_emb) %*% t(l2_norm(concept_mat))
colnames(scores) <- names(concept_probes)

as_tibble(scores) |>
  bind_cols(poem_lines |> select(line_num, text)) |>
  pivot_longer(
    cols      = all_of(names(concept_probes)),
    names_to  = "concept",
    values_to = "score"
  ) |>
  mutate(
    concept    = factor(concept, levels = names(concept_probes)),
    short_text = fct_inorder(paste0(line_num, ": ", str_trunc(text, 28)))
  ) |>
  ggplot(aes(x = short_text, y = score, fill = concept)) +
  geom_col() +
  facet_wrap(~ concept, ncol = 1, scales = "free_y") +
  theme_minimal(base_size = 8) +
  theme(
    axis.text.x     = element_text(angle = 55, hjust = 1),
    legend.position = "none",
    panel.spacing   = unit(0.8, "lines")
  ) +
  labs(
    title = glue::glue("Line scores along concept directions — {poem_title}"),
    x     = NULL,
    y     = "Cosine similarity to concept"
  )
```

---

## Whole-Poem Topic Scores

Each keyword is embedded and every line is scored against it via cosine
similarity. The whole-poem score is the mean across lines, with an alternative
word-count-weighted mean for poems where line length varies substantially.
The result is a compact topic-score vector that can be saved and stacked with
scores from other poems for clustering or comparison.

```{r}
#| label: topic-scores
#| cache: true
#| dependson: ["embeddings", "config"]

kw_result <- openai::create_embedding(
  model = "text-embedding-3-small",
  input = topic_keywords
)
kw_mat <- do.call(rbind, kw_result$data$embedding)

line_topic_scores <- l2_norm(line_emb) %*% t(l2_norm(kw_mat))
colnames(line_topic_scores) <- topic_keywords

word_counts <- str_count(poem_lines$text, "\\S+")

poem_topic_vector <- tibble(
  poem        = poem_title,
  author      = poem_author,
  topic       = topic_keywords,
  mean_score  = round(colMeans(line_topic_scores), 4),
  wtd_score   = round(apply(line_topic_scores, 2,
                            \(x) weighted.mean(x, w = word_counts)), 4)
)

poem_topic_vector
```

```{r}
#| label: topic-scores-plot
#| fig-width: 7
#| fig-height: 5
#| dependson: "topic-scores"

# Line-level detail with whole-poem mean as reference
as_tibble(line_topic_scores) |>
  bind_cols(poem_lines |> select(line_num, text)) |>
  pivot_longer(all_of(topic_keywords), names_to = "topic", values_to = "score") |>
  mutate(
    topic      = factor(topic, levels = topic_keywords),
    short_text = fct_inorder(paste0(line_num, ": ", str_trunc(text, 28)))
  ) |>
  ggplot(aes(x = short_text, y = score)) +
  geom_col(fill = "steelblue", alpha = 0.75) +
  geom_hline(
    data = poem_topic_vector |> mutate(topic = factor(topic, levels = topic_keywords)),
    aes(yintercept = mean_score),
    linetype = "dashed", color = "firebrick", linewidth = 0.7
  ) +
  facet_wrap(~ topic, ncol = 1, scales = "free_y") +
  theme_minimal(base_size = 8) +
  theme(
    axis.text.x     = element_text(angle = 55, hjust = 1),
    legend.position = "none",
    panel.spacing   = unit(0.8, "lines")
  ) +
  labs(
    title    = glue::glue("Topic scores by line — {poem_title}"),
    subtitle = "Dashed line = whole-poem mean",
    x        = NULL,
    y        = "Cosine similarity to keyword"
  )
```

```{r}
#| label: save-topic-vector
#| eval: false

# Uncomment and run manually to accumulate scores across poems.
# Each poem appends one row per topic to the CSV.
readr::write_csv(
  poem_topic_vector,
  "poem_topic_scores.csv",
  append = file.exists("poem_topic_scores.csv")
)
```
